---
title: "Lab 9b - Solution code"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=F}
library(tidyverse)
```



# One-way ANOVA

```{r}
algorithm <- read.csv("algorithm.csv", header=T, stringsAsFactors = T)

```

## Dot plot for each algorithm

### Algorithm A

```{r}
p <- ggplot(algorithm, aes(x= Algorithm.A)) + 
  geom_dotplot(binwidth=0.5, col="red", fill="red")  
p <- p+ theme_classic() + labs (x="Algorithm A")
p <- p + theme(text = element_text(size = 20)) 
p
```

### Algorithm B

```{r}
p <- ggplot(algorithm, aes(x= Algorithm.B)) + 
  geom_dotplot(binwidth=0.5, col="red", fill="red")  
p <- p+ theme_classic() + labs (x="Algorithm B")
p <- p + theme(text = element_text(size = 20)) 
p
```

### Algorithm C

```{r}
p <- ggplot(algorithm, aes(x= Algorithm.C)) + 
  geom_dotplot(binwidth=0.5, col="red", fill="red")  
p <- p+ theme_classic() + labs (x="Algorithm C")
p <- p + theme(text = element_text(size = 20)) 
p
```


### Algorithm D

```{r}
p <- ggplot(algorithm, aes(x= Algorithm.D)) + 
  geom_dotplot(binwidth=0.5, col="red", fill="red")  
p <- p+ theme_classic() + labs (x="Algorithm D")
p <- p + theme(text = element_text(size = 20)) 
p
```


## Data preparation

Converting the dataset to the right format - 2 columns, one for the algorithm and one for the time.

```{r}
# creating mini data frames, one for each algorithm, 
# each one containscone column with the name of the algorithm and one column with the time
algoA <- data.frame(time = algorithm$Algorithm.A, algorithm= "A")
algoB <- data.frame(time = algorithm$Algorithm.B, algorithm= "B")
algoC <- data.frame(time = algorithm$Algorithm.C, algorithm= "C")
algoD <- data.frame(time = algorithm$Algorithm.D, algorithm= "D")
# putting the 4 mini data frames together
dataAlgo <- rbind(algoA, algoB, algoC,algoD)

```

## One-way ANOVA test 

Applying the one-way ANOVA test.
 
H0: There is no difference in mean values i.e. µA = µB = µC = µD.

H1: at least two means are different.

 
```{r}
anova <- aov( time ~ algorithm, data = dataAlgo)
summary(anova)
```


The test indicates that there is no significant difference in means.

### Checking the validity of the test

The residuals need to be checked. The test is only valid if the residuals are normally distributed.

#### Q-Q plot

```{r}
# first construct a data frame with the rediduals. Then use it in the plot
anovaFrame <- data.frame(residuals = anova$residuals )
p <- ggplot(anovaFrame, aes(sample = residuals)) 
p <- p + stat_qq(size=2) + stat_qq_line( alpha = 0.7, color = "red", linetype = "dashed")
p <- p +  theme_classic()
p <- p + theme(text = element_text(size = 20))
p
```

The points are close to the theoretical line. It is reasonable to assume the distribution is normal.

#### Shapiro-Wilk test of normality

```{r}
shapiro.test(anova$residuals)
```


The p-value   (0.4428) is > 0.05 so it is reasonable to assume the distribution is normal.


## Sample sizes

### Sample sizes for t-test

To get the number of samples required for a power of 0.8, with a difference of 8 units, an estimated standard deviation of 10 and a significance level of 0.05.

```{r}
a <- power.t.test(power=0.8,delta=8,sd=10,sig.level=0.05,type="two.sample")
a
```

# Exercises {-}


#  Exercise 1



```{r}
methods <- read.csv("methods.csv", header=T,stringsAsFactors=T)
```

```{r}
# creating mini data frames, one for each algorithm, 
# each one contains cone column with the number of documents and another with the method.
methodA <- data.frame(docNumber = methods$Method.A, method= "A")
methodB <- data.frame(docNumber = methods$Method.B, method= "B")
methodC <- data.frame(docNumber = methods$Method.C, method= "C")
# putting the 4 mini data frames together
dataMethod <- rbind(methodA, methodB, methodC)
```

## One-way ANOVA test 

Applying the one-way ANOVA test.
 
H0: There is no difference in mean values i.e. µA = µB = µC .

H1: at least two means are different.

 
```{r}
anova <- aov( docNumber ~ method, data = dataMethod)
summary(anova)
```
p-value > 0.05. Do not reject  . There is not enough evidence of any differences between the mean numbers of relevant documents retrieved by the three methods.

The valididy of the test  needs to be checked.

### Checking whether the residuals are normally distributed

Checking normal distribution of residuals using a Q-Q plot.

```{r}
anovaFrame <- data.frame(residuals = anova$residuals )
p <- ggplot(anovaFrame, aes(sample = residuals)) 
p <- p + stat_qq() 
p <- p + stat_qq_line( )
p
```

The points are close to the line suggesting it is reasonable to assume that the distribution is normal.

Checking normal distribution of residuals using the Shapiro-Wilk test

```{r}
shapiro.test(anovaFrame$residuals)
```

A p-value > 0.05 indicates that it is reasonable to assume that the residuals are normally distributed.


# Exercise 2



```{r}
tool <- read.csv("tool.csv", header=T,stringsAsFactors=T)
```

```{r}
# creating mini data frames, one for each algorithm, 
# each one contains cone column with the number of documents and another with the method.
toolA <- data.frame(percentage = tool$Tool.A, tool = "A")
toolB <- data.frame(percentage = tool$Tool.B, tool= "B")
toolC <- data.frame(percentage = tool$Tool.C, tool= "C")
toolD <- data.frame(percentage = tool$Tool.D, tool= "D")
# putting the 4 mini data frames together
dataTool <- rbind(toolA, toolB, toolC, toolD)
```


H0: There is no difference in mean values i.e. µA = µB = µC = µD.

H1: at least two means are different.

```{r}
anova <- aov( percentage ~ tool, data = dataTool)
summary(anova)
```

The p-value (5.78e-06) indicates that there are differences in means. 

The valididy of the test  needs to be checked.

### Checking whether the residuals are normally distributed

Checking normal distribution of residuals using a Q-Q plot.

```{r}
anovaFrame <- data.frame(residuals = anova$residuals )
p <- ggplot(anovaFrame, aes(sample = residuals)) 
p <- p + stat_qq() 
p <- p + stat_qq_line( )
p
```

The points are close to the line, although one of the points (lower left) may be a cause for concern.

Checking normal distribution of residuals using the Shapiro-Wilk test

```{r}
shapiro.test(anovaFrame$residuals)
```

A p-value > 0.05 indicates that it is reasonable to assume that the residuals are normally distributed.

# Exercise 3

Sample sizes for different powers with 5% significance level and a standard deviation of 10 and a difference of 8.

```{r}

a <- power.t.test(power=0.8,delta=8,sd=10,sig.level=0.05,type="two.sample")
b<- power.t.test(power=0.85,delta=8,sd=10,sig.level=0.05,type="two.sample")
c <- power.t.test(power=0.9,delta=8,sd=10,sig.level=0.05,type="two.sample")
d <- power.t.test(power=0.95,delta=8,sd=10,sig.level=0.05,type="two.sample")
e <- power.t.test(power=0.99,delta=8,sd=10,sig.level=0.05,type="two.sample")
f <- power.t.test(power=0.999,delta=8,sd=10,sig.level=0.05,type="two.sample")

## Displaying the results

cat(   "",
       "n = ", ceiling(a$n), "units = ", a$delta, "power = ",  a$power, "significance level = ",  a$sig.level, "\n", 
       "n = ", ceiling(b$n), "units = ", b$delta, "power = ",  b$power, "significance level = ",  b$sig.level, "\n",
       "n = ", ceiling(c$n), "units = ", c$delta, "power = ",  c$power,  "significance level = ",  c$sig.level,  "\n",
       "n = ", ceiling(d$n), "units = ", d$delta, "power = ",  d$power, "significance level = ",  d$sig.level,  "\n",
       "n = ", ceiling(e$n), "units = ", e$delta, "power = ",  e$power, "significance level = ",  e$sig.level,  "\n",
       "n = ", ceiling(f$n), "units = ", f$delta, "power = ", f$power,  "significance level = ",  f$sig.level, sep = "\t")

```

Displaying the results graphically.

```{r}
powerVal <- c(a$power,b$power, c$power, d$power, e$power, f$power)
sampleSize <- c(a$n, b$n, c$n, d$n, e$n, f$n)
data <- data.frame(power = powerVal, sampleSeize = sampleSize)
p <- ggplot(data, aes(x=power, y=sampleSize))
p <- p+ geom_point(colour="red", size=3) + theme_classic() +labs(x= "Power", y= "Sample size (n)")
p <- p  + theme(text = element_text(size = 20))
p
```

## Exercise 4

### Section a

```{r}
power.t.test(power=0.7,delta=10,sd=25,sig.level=0.05,type="two.sample", alternative ="two.sided")
power.t.test(power=0.8,delta=10,sd=25,sig.level=0.05,type="two.sample", alternative ="two.sided")
power.t.test(power=0.9,delta=10,sd=25,sig.level=0.05,type="two.sample", alternative ="two.sided")
power.t.test(power=0.95,delta=10,sd=25,sig.level=0.05,type="two.sample", alternative ="two.sided")

power.t.test(power=0.7,delta=15,sd=25,sig.level=0.05,type="two.sample", alternative ="two.sided")
power.t.test(power=0.8,delta=15,sd=25,sig.level=0.05,type="two.sample", alternative ="two.sided")
power.t.test(power=0.9,delta=15,sd=25,sig.level=0.05,type="two.sample", alternative ="two.sided")
power.t.test(power=0.95,delta=15,sd=25,sig.level=0.05,type="two.sample", alternative ="two.sided")

power.t.test(power=0.7,delta=20,sd=25,sig.level=0.05,type="two.sample", alternative ="two.sided")
power.t.test(power=0.8,delta=20,sd=25,sig.level=0.05,type="two.sample", alternative ="two.sided")
power.t.test(power=0.9,delta=20,sd=25,sig.level=0.05,type="two.sample", alternative ="two.sided")
power.t.test(power=0.95,delta=20,sd=25,sig.level=0.05,type="two.sample", alternative ="two.sided")
```

### Section b


```{r}
power.t.test(power=0.7,delta=10,sd=25,sig.level=0.01,type="two.sample", alternative ="two.sided")
power.t.test(power=0.8,delta=10,sd=25,sig.level=0.01,type="two.sample", alternative ="two.sided")
power.t.test(power=0.9,delta=10,sd=25,sig.level=0.01,type="two.sample", alternative ="two.sided")
power.t.test(power=0.95,delta=10,sd=25,sig.level=0.01,type="two.sample", alternative ="two.sided")

power.t.test(power=0.7,delta=15,sd=25,sig.level=0.01,type="two.sample", alternative ="two.sided")
power.t.test(power=0.8,delta=15,sd=25,sig.level=0.01,type="two.sample", alternative ="two.sided")
power.t.test(power=0.9,delta=15,sd=25,sig.level=0.01,type="two.sample", alternative ="two.sided")
power.t.test(power=0.95,delta=15,sd=25,sig.level=0.01,type="two.sample", alternative ="two.sided")

power.t.test(power=0.7,delta=20,sd=25,sig.level=0.01,type="two.sample", alternative ="two.sided")
power.t.test(power=0.8,delta=20,sd=25,sig.level=0.01,type="two.sample", alternative ="two.sided")
power.t.test(power=0.9,delta=20,sd=25,sig.level=0.01,type="two.sample", alternative ="two.sided")
power.t.test(power=0.95,delta=20,sd=25,sig.level=0.01,type="two.sample", alternative ="two.sided")
```

### Section c

```{r}
power.t.test(power=0.7,delta=10,sd=25,sig.level=0.01,type="two.sample", alternative ="one.sided")
power.t.test(power=0.8,delta=10,sd=25,sig.level=0.01,type="two.sample", alternative ="one.sided")
power.t.test(power=0.9,delta=10,sd=25,sig.level=0.01,type="two.sample", alternative ="one.sided")
power.t.test(power=0.95,delta=10,sd=25,sig.level=0.01,type="two.sample", alternative ="one.sided")

power.t.test(power=0.7,delta=15,sd=25,sig.level=0.01,type="two.sample", alternative ="one.sided")
power.t.test(power=0.8,delta=15,sd=25,sig.level=0.01,type="two.sample", alternative ="one.sided")
power.t.test(power=0.9,delta=15,sd=25,sig.level=0.01,type="two.sample", alternative ="one.sided")
power.t.test(power=0.95,delta=15,sd=25,sig.level=0.01,type="two.sample", alternative ="one.sided")

power.t.test(power=0.7,delta=20,sd=25,sig.level=0.01,type="two.sample", alternative ="one.sided")
power.t.test(power=0.8,delta=20,sd=25,sig.level=0.01,type="two.sample", alternative ="one.sided")
power.t.test(power=0.9,delta=20,sd=25,sig.level=0.01,type="two.sample", alternative ="one.sided")
power.t.test(power=0.95,delta=20,sd=25,sig.level=0.01,type="two.sample", alternative ="one.sided")
```

