---
title: "Lab4ggplot"
author: "Dulla"
date: "2024-04-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tibble)
library(ggplot2)
library(tidyverse)
```

```{r}
consumptionNoOil<- read.csv('consumptionNoOil.csv', header = T, stringsAsFactors = T)
projectTeam <- read.csv('projectTeam.csv', header = T, stringsAsFactors = T)
importsExportsAmerica <- read.csv('importsExportsAmerica.csv', header = T, stringsAsFactors = T)
provenOilReserveWEurope <- read.csv('provenOilReserveWEurope.csv', header = T, stringsAsFactors = T)
channels<-read.csv('channels.csv', header = T, stringsAsFactors = T) 
uni <- read.csv('universities.csv', header = T, stringsAsFactors = T) 
```

```{r}
p <- ggplot(importsExportsAmerica, aes(crude.imports,crude.exports,
label=country))
p <- p+geom_text()
p
```

```{r}
p <- ggplot(importsExportsAmerica, aes(crude.imports,crude.exports,
label=country))
p <- p+geom_area()
p
```


```{r}
p <- ggplot(importsExportsAmerica, aes(crude.imports,crude.exports,
label=country))
p <- p+geom_polygon()
p
```

```{r}
p <- ggplot(importsExportsAmerica, aes(crude.imports,crude.exports,
label=country, color=country))
p <- p+geom_tile()
p
```

```{r}
p <- ggplot(projectTeam, aes(site, fill = experience))
p <- p + geom_bar(position = "stack")
p
```

```{r}
p <- ggplot(projectTeam, aes(site, fill = experience))
p <- p + geom_bar(position = "dodge")
p
```
```{r}
p <- ggplot(projectTeam, aes(site, fill = experience))
p <- p + geom_bar(position = "fill")
p
```

```{r}
p <- ggplot(projectTeam, aes(site, fill = experience))
p <- p + geom_bar(position = "identity")
p
```
```{r}
p <- ggplot(consumptionNoOil, aes(Year,MTOE, colour=Source))
p <- p+geom_point()
p
```

```{r}
p <- ggplot(consumptionNoOil, aes(Year,MTOE, colour=Source))
p <- p + xlim(2010,2035)
p <- p+geom_point()
p
```
```{r}
p <- ggplot(consumptionNoOil, aes(Year,MTOE, colour=Source))
p <- p+geom_point()
p<- p+ labs(x="Year", y="Millions of tons of oil equivalent",title ="Energy consumption",subtitle = "Oil is excluded")
p
```

```{r}
p <- ggplot(consumptionNoOil, aes(Year,MTOE, colour=Source))
p <- p+geom_point()
p<- p+ labs(x="Year","Millions of tons of oil equivalent",
title ="Energy consumption by source",
subtitle = "Oil is excluded",
colour = "Energy source")
p
```
```{r}
colours <-c(coal="black", gas="yellow2", renewables="green")
p <- ggplot(consumptionNoOil, aes(Year, MTOE, colour=Source))
p <- p +geom_point()
p<-p+labs(x= "Year",y= "Millions of tons of oil equivalent",title = "Energy consumed",colour="Energy source")
p<- p + scale_colour_manual(values=colours)
p
```
```{r}
p <- ggplot(consumptionNoOil, aes(Year,MTOE, colour=Source))
p <- p+geom_point()
p <- p + geom_vline(aes(xintercept = Year), linetype = "longdash",
colour="green")
p
```

```{r}
p <- ggplot(consumptionNoOil, aes(Year,MTOE, colour=Source))
p <- p+geom_point()
p <- p + geom_vline(aes(xintercept = 2015), linetype = "longdash",
colour="green")
p
```

```{r}
p <- ggplot(consumptionNoOil, aes(Year,MTOE))
p <- p+geom_point()
p <- p + geom_hline(aes(yintercept = 100), linetype = "dotdash",
colour="green")
p
```
```{r}
p <- ggplot(consumptionNoOil, aes(Year,MTOE, colour=Source))
p <- p+geom_point()
p <- p + geom_hline(aes(yintercept = 100), linetype = "dotted",
colour="darkgreen")
p <- p + geom_vline(aes(xintercept = 2015), linetype = "twodash",
colour="blue")
p
```
# counting the number of staff per site.
```{r}
pieData <- projectTeam |> count(site, name="Number")
names(pieData) <- c("site","numberStaff")
```

# order data according to the site (important for placing labels)
```{r}
pieData <- arrange(pieData,desc(site))
```

# create new column with position for label
```{r}
pieData <- pieData |>
mutate(positionLabel = cumsum(numberStaff) - 0.5* numberStaff)
```

# creating plot
```{r}
p <- ggplot(pieData, aes(x="", y= numberStaff, fill = site))
p <- p + geom_bar(width = 1, stat="identity")
p <- p + coord_polar("y", start=0)
```

# adding text labels
```{r}
p <- p + geom_text(aes(y = positionLabel, label = numberStaff))
```

# Changing the colour scheme
```{r}
p <- p + scale_fill_manual(values = c("red","blue", "lightblue"))
```

# removing outer "ring"
```{r}
p <- p + theme_void()
p
```
```{r}
p <- ggplot(data=ProvenOilReservesZerosT, aes(MTBarrels, Denmark))
p <- p + geom_point()
p <- p + labs(title="Yearly proven oil vs Denmark", subtitle = "Denmark", x= "Year of Production", y="Number of Barrels ")
p <- p + geom_hline(aes(yintercept= 0.75), linetype= "twodash")
p
```
```{r}
p <- ggplot(data = ProvenOilReservesZerosT, aes(MTBarrels, Italy))
p <- p + geom_line(size=2, color= "orange")
p <- p + labs(title="This the title", x= "This is the xaxis", y="This is the y axis")
p
```

Using the channels dataset, plot length against area using a scatterplot. Use size to
indicate “n” (the size of the stimuli).
```{r}
l <- ggplot(data = channels, aes(length, area, size = n))
l <- l + geom_point()
l <- l + 
l
              
```

Repeat the exercise above, but this time change the y axis to go from 0 to 5, so that it has the same scale as the x axis. Compare your plot with your previous plot, paying particular attention to which one is more effective to compare length against area. Ensure appropriate titles and labels.
```{r}
l <- ggplot(data = channels, aes(length, area, size = n))
l <- l + geom_point()
p <- p + labs(x="length", y="area" , title="Perception - length vs. area", size= "Size of stimuli")
l <- l + ylim(0,5)
l
```
```{r}
p <- ggplot(uni, aes(status, fill = school))
p <- p + geom_bar(position = "fill")
p
```
```{r}
uni2 <- uni |>filter(status %in% c('student', 'both'))
```

# Counting the number of rows per school
```{r}
pieData <- uni2 |> count(school, name="numberStudents")
```

#Ordering data according to the school name (important for placing labels later on)
```{r}
pieData <- arrange(pieData,desc(school))
```

# Creating a new column called positionLabel with position for label
```{r}
pieData <-  mutate(pieData, positionLabel = cumsum(pieData$numberStudents) - 0.5*pieData$numberStudents)
```

# creating pie chart
```{r}
p <- ggplot(pieData,  aes(x="", y= numberStudents, fill = school))
p <- p + geom_bar(width = 1, stat="identity")
p <- p + coord_polar("y", start=0)
```

# Adding text labels
```{r}
p <- p + geom_text(aes(y = positionLabel, label = numberStudents))
```

# Changing the colour scheme
```{r}
p <- p +   scale_fill_manual(values = c("wheat", "lightblue", "cornflowerblue", "blue1")) 
```

# removing outer "ring"
```{r}
p <- p + theme_void()
p
```

