---
title: "Lab6_Data_Analysis"
author: "Dulla"
date: "2024-04-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
 rm(list = ls())
library(tidyverse)
```



```{r}
Consumption <- read.csv("consumptionTransposed.csv", header = T, stringsAsFactors = T)
```

```{r}
summary(Consumption)
str(Consumption)
class(Consumption)
head(Consumption)
tail(Consumption)
```


```{r}
Consumption |>  select(Oil) |> summary()
```

```{r}
Consumption |> pull(Gas)
```

```{r}
Consumption |> pull(Gas) |> mean()
```

```{r}
browserdata <- read.csv("browserdata.csv",header = T, stringsAsFactors = T)
summary(browserdata)
```
One-way tables
```{r}
counts <- browserdata |> pull(Browser) |> table()
counts
```

```{r}
prop.table(counts)  #each row adds up to 1
```

Two-way tables
```{r}
freqBrowser <- xtabs(~ Browser+Membership, data=browserdata)
freqBrowser
```
```{r}
prop.table(freqBrowser,2) # each column adds up to 1
```
# convert data to dataframe first
```{r}
e <- as.data.frame(browserdata)
e
```

For a simple bar plot
```{r}
p <- ggplot(e, aes(x=Membership,fill=Browser))
p <- p + geom_bar(aes(y=Use), stat="identity")
p <- p + labs(title = "Browser use by membership type",
x = "Membership type", y = "Frequency count")
p <- p + scale_fill_brewer(palette="Set1")
p <- p + theme_classic()
p
```

Similarly, the bar plot could show each browser data in a different column.

```{r}
p <- ggplot(e, aes(x=Membership, y=Use, fill=Browser) )
p <- p + geom_bar(stat="identity", position=position_dodge())
p <- p + labs(title = "Browser use by membership type", x =
"membership type", y="Frequency count")
p <- p + scale_fill_brewer(palette="Dark2")
p <- p + theme_classic()
p
```
Create a histogram of the daily time spent online as follows:
```{r}
p <- ggplot(browserdata, aes(x= MinsPerDay))
p <- p + geom_histogram(aes(y=..density..), binwidth=20 )
p
```
Adding colour, titles, labels, and a line
```{r}
p <- ggplot(browserdata, aes(x= MinsPerDay))
p <- p + geom_histogram(aes(y=..density..),col="red", fill="red", binwidth=10)
p <- p + labs(x="Average time spent online (mins/day)",y="Density")
p <- p + geom_density(col="blue")
p <- p + theme_classic()
p
```

The following code produces a dot plot for column Use, which shows the distribution of values.
```{r}
p <- ggplot(browserdata, aes(x= Use))
p <- p + geom_dotplot(binwidth=0.05)
p
```
The following produces a nicer-looking dot plot.
```{r}
p <- ggplot(browserdata, aes(x= Use))
p <- p + geom_dotplot(col="red", fill="red" , binwidth=0.05)
p <- p + labs(x="Use level", y="")
p <- p + theme_classic()
p
```

##Box plots
The following code produces a boxplot for the average daily time online.
```{r}
p <- ggplot(browserdata, aes(y= MinsPerDay))
p <- p + geom_boxplot()
p
```


Adding colour, titles, etc.
```{r}
p <- ggplot(browserdata, aes(y= MinsPerDay))
p <- p + geom_boxplot(col="blue", fill="lightblue")
p <- p + labs(title="User average minutes/day online",
x="", y="")
p <- p + theme_classic()
p
```

```{r}
p <- ggplot(browserdata, aes(x=Browser, y= MinsPerDay))
p <- p + geom_boxplot(aes(col=factor(Browser)) )
p <- p + labs(title="User average minutes/day online by browser",
x="",y="Average minutes per day")
p <- p + theme_classic()
p
```

```{r}
p <- ggplot(data.frame(x=c(-pi,pi)),aes(x=x))
p <- p + stat_function(fun=cos)
p
```

```{r}
mysquare <- function(xval) {xval*xval}
xlimits <- c(-pi,pi) ## pi = 3.14...
xlimits
```

```{r}
p <- ggplot(data.frame(xlimits),aes(x=xlimits))
p <- p + stat_function(fun=cos) 
p
```
Try replacing geom_histogram with geom_freqpoly and geom_density.
```{r}
normal.sampled <- rnorm(1000)
p <- ggplot(NULL,aes(x=normal.sampled,y=..density..))
p <- p + geom_histogram(binwidth = 0.1)
p
```

```{r}
normal.m1.s1 <- rnorm(1000,mean=1,sd=1)
normal.m2.s3 <- rnorm(1000,mean=2,sd=3)
norm.compare <- data.frame(normal.m1.s1,normal.m2.s3)
bplot.data <- stack(norm.compare)
p <- ggplot(bplot.data,aes(x=ind,y=values,fill=ind))
p <- p + geom_boxplot()
p
```




